---
title: "Endometrisis_analysis"
author: "Iza"
date: "2026-03-10"
output: html_document
---


STEP 1: Install and Load Required Packages
# We need, besides MetaVolcanoR:
- GEOquery    : download data directly from GEO
- limma       : differential expression for microarray/RNA-seq
- dplyr/tidyr : data wrangling
- ggplot2     : plotting
- BiocManager : install Bioconductor packages
```{r}
pkgs_bioc <- c("GEOquery", "limma", "DESeq2", "edgeR", "Biobase")
pkgs_cran <- c("dplyr", "tidyr", "ggplot2", "MetaVolcanoR", "stringr", "readr")

for (pkg in pkgs_bioc) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message("Installing ", pkg, " ...")
    BiocManager::install(pkg, ask = FALSE)
  }
}

for (pkg in pkgs_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message("Installing ", pkg, " ...")
    install.packages(pkg)
  }
}
```

Load all packages
```{r}
library(GEOquery)
library(limma)
library(DESeq2)
library(edgeR)
library(Biobase)
library(dplyr)
library(tidyr)
library(ggplot2)
library(MetaVolcanoR)
library(stringr)

```

Since there is raw GEO data download setting higher timeouts
```{r}
# Set global options
options(timeout = 300)  # longer timeout for GEO downloads
set.seed(42)

# Output directory
dir.create("results", showWarnings = FALSE)
dir.create("results/DE",    showWarnings = FALSE)
dir.create("results/plots", showWarnings = FALSE)
```

# ==============================
# STEP 2: Define Sample Annotations
# =============================
# We explicitly define which GSM IDs are cases and which are controls this matches what is available for each study on GEO.
[EDIT] Add here somethign about endometriosis and why to do this.
```{r}
# Step 1 — Sample annotations (updated)
sample_info <- list(

  GSE25628 = list(
    cases    = c("GSM629719", "GSM629721", "GSM629723", "GSM629726",
                 "GSM629728", "GSM629730", "GSM629732", "GSM629740"),
    controls = c("GSM629722", "GSM629724", "GSM629725", "GSM629727",
                 "GSM629729", "GSM629735"),
    type     = "microarray"
  ),

  GSE145701 = list(
    cases    = c("GSM4331218", "GSM4331222", "GSM4331226", "GSM4331230",
                 "GSM4331234", "GSM4331238", "GSM4331242", "GSM4331246"),
    controls = c("GSM4331202", "GSM4331206", "GSM4331210", "GSM4331214"),
    type     = "microarray"
  ),

  GSE120103 = list(
    cases    = c(paste0("GSM", 3393500:3393508),
                 paste0("GSM", 3393518:3393526)),
    controls = c(paste0("GSM", 3393491:3393499),
                 paste0("GSM", 3393509:3393517)),
    type     = "microarray"
  ),
  GSE25628 = list(
    cases    = c("GSM629719", "GSM629721", "GSM629723", "GSM629726",
                 "GSM629728", "GSM629730", "GSM629732", "GSM629740"),
    controls = c("GSM629722", "GSM629724", "GSM629725", "GSM629727",
                 "GSM629729", "GSM629735"),
    type     = "microarray"
  ),
  GSE145701 = list(
    cases    = c("GSM4331218", "GSM4331222", "GSM4331226", "GSM4331230",
                 "GSM4331234", "GSM4331238", "GSM4331242", "GSM4331246"),
    controls = c("GSM4331202", "GSM4331206", "GSM4331210", "GSM4331214"),
    type     = "microarray"
  ),
  GSE120103 = list(
    cases    = c(paste0("GSM", 3393500:3393508), paste0("GSM", 3393518:3393526)),
    controls = c(paste0("GSM", 3393491:3393499), paste0("GSM", 3393509:3393517)),
    type     = "microarray"
  ),
  # --- new datasets ---
  GSE23339 = list(
    cases    = c("GSM572747", "GSM572748", "GSM572749", "GSM572750", "GSM572751",
                 "GSM572758", "GSM572760", "GSM572763", "GSM572765"),
    controls = c("GSM572752", "GSM572753", "GSM572754", "GSM572755", "GSM572756",
                 "GSM572757", "GSM572761", "GSM572762", "GSM572764"),
    type     = "microarray"
  ),
  GSE17504 = list(
    cases    = paste0("GSM", 436440:436444),
    controls = paste0("GSM", 436450:436454),
    type     = "microarray"
  ),
  GSE7846 = list(
    cases    = paste0("GSM", 190395:190399),
    controls = paste0("GSM", 190400:190404),
    type     = "microarray"
  ),
  GSE6364 = list(
    cases    = paste0("GSM", 150190:150195),
    controls = paste0("GSM", 150196:150201),
    type     = "microarray"
  ),
  GSE7305 = list(
    cases    = paste0("GSM", 175766:175775),
    controls = paste0("GSM", 175776:175785),
    type     = "microarray"
  )
)

```

Download GEO datasets
# GEOquery::getGEO() downloads the series matrix file from NCBI GEO.
```{r}

destdir <- normalizePath("results", mustWork = FALSE)

geo_data <- list()
for (gse_id in names(sample_info)) {
  cache_file <- file.path(destdir, paste0(gse_id, "_gse.rds"))

  if (file.exists(cache_file)) {
    message(gse_id, ": loading from cache...")
    geo_data[[gse_id]] <- readRDS(cache_file)
  } else {
    message(gse_id, ": downloading from GEO...")
    gse <- tryCatch(
      getGEO(gse_id, GSEMatrix = TRUE, getGPL = TRUE, destdir = destdir),
      error = function(e) { message("  ERROR: ", e$message); NULL }
    )
    if (!is.null(gse)) {
      geo_data[[gse_id]] <- gse
      saveRDS(gse, cache_file)
      message(gse_id, ": download complete.")
    }
  }
}
```


# STEP 4: Differential Expression per Dataset (GEO2R approach)
# ==============================================================================
# This reproduces what NCBI GEO2R does interactively:
#   1. Download the series matrix (already done above)
#   2. Subset to the user-defined case/control samples
#   3. Fit a linear model (limma) 
#   4. Extract log2 fold-change, p-values, and adjusted p-values
#   5. Remember to save the variance reuslts form limma

```{r}
run_limma_de <- function(eset, group, dataset_id) {
  message("  Running limma DE for ", dataset_id)
  
  design    <- model.matrix(~ group)
  exprs_mat <- exprs(eset)
  exprs_mat <- exprs_mat[complete.cases(exprs_mat), ]
  
  fit <- lmFit(exprs_mat, design)
  fit <- eBayes(fit)
  
  tt <- topTable(fit, coef = 2, number = Inf,
                 adjust.method = "BH", sort.by = "none",
                 confint = TRUE)
  
  fdata    <- fData(eset)
  gene_col <- intersect(
    c("Gene Symbol", "GENE_SYMBOL", "gene_symbol", "Symbol", "SYMBOL",
      "Gene.Symbol", "GeneSymbol", "gene_assignment", "mrna_assignment"),
    colnames(fdata)
  )[1]
  
  symbols <- if (!is.na(gene_col)) {
    as.character(fdata[rownames(tt), gene_col, drop = TRUE])
  } else {
    rownames(tt)
  }
  
  data.frame(
    probe_id = rownames(tt),
    symbol   = as.character(symbols),
    log2FC   = tt$logFC,
    CI.L     = tt$CI.L,
    CI.R     = tt$CI.R,
    pvalue   = tt$P.Value,
    adj_pval = tt$adj.P.Val,
    dataset  = dataset_id,
    stringsAsFactors = FALSE
  )
}

de_results <- list()

make_group <- function(eset, cases, controls) {
  samples <- sampleNames(eset)
  group   <- ifelse(samples %in% cases, "case", "control")
  factor(group, levels = c("control", "case"))  # control is reference level
}

for (gse_id in names(sample_info)) {
  info     <- sample_info[[gse_id]]
  eset     <- if (is.list(geo_data[[gse_id]])) geo_data[[gse_id]][[1]] else geo_data[[gse_id]]
  all_gsm  <- c(info$cases, info$controls)
  eset_sub <- eset[, sampleNames(eset) %in% all_gsm]
  
  found          <- sampleNames(eset_sub)
  cases_found    <- sum(found %in% info$cases)
  controls_found <- sum(found %in% info$controls)
  message(gse_id, ": cases=", cases_found, " controls=", controls_found)
  
  if (cases_found < 2 || controls_found < 2) {
    message("  WARNING: insufficient samples — skipping"); next
  }
  
  group <- make_group(eset_sub, info$cases, info$controls)
  de_results[[gse_id]] <- run_limma_de(eset_sub, group, gse_id)
  
  write.csv(de_results[[gse_id]],
            file.path("results/DE", paste0(gse_id, "_DE_results.csv")),
            row.names = FALSE)
  message(gse_id, ": DE complete — ", nrow(de_results[[gse_id]]), " probes")
}
```

# ==============================================================================
# STEP 5: STANDARIZE Gene names for running
[EDIT] explain here why in detail
```{r}
clean_gene_symbols <- function(df) {
  
  df %>%
    mutate(
      symbol = case_when(
        # "NM_xxx // GENESYMBOL // description" → field 2
        grepl(" // ",  symbol) ~ str_trim(str_split_fixed(symbol, " // ",  3)[, 2]),
        # "GENESYMBOL /// GENESYMBOL2"          → field 1
        grepl(" /// ", symbol) ~ str_trim(str_split_fixed(symbol, " /// ", 2)[, 1]),
        TRUE ~ str_trim(symbol)
      ),
      symbol = str_to_upper(symbol)
    ) %>%
    filter(
      !is.na(symbol) & symbol != "" & symbol != "---" & symbol != "NA" &
      !grepl("^[0-9]",      symbol) &
      !grepl("^LOC[0-9]",   symbol) &
      !grepl("^\\(\\+\\)",  symbol) &
      !grepl("^ASHGV",      symbol) &
      !grepl("^ENST",       symbol) &
      !grepl("^ENSG",       symbol) &
      !grepl("^A_23_P",     symbol) &   # Agilent probe IDs
      !grepl("^A_24_P",     symbol)
    ) %>%
    group_by(symbol) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup()
}

de_results <- lapply(de_results, clean_gene_symbols)

# Rebuild mv_input
mv_input <- lapply(de_results, function(df) as.data.frame(df))

# Sanity check
cat("=== mv_input status ===\n")
lapply(names(mv_input), function(nm) {
  df <- mv_input[[nm]]
  cat(nm, "| genes:", nrow(df),
      "| cols:", paste(colnames(df), collapse = ", "),
      "| sample:", paste(head(df$symbol, 3), collapse = ", "), "\n")
})

# Overlap check
gene_lists <- lapply(mv_input, `[[`, "symbol")
cat("\n=== Overlap ===\n")
nms <- names(gene_lists)
for (i in 1:(length(nms)-1))
  for (j in (i+1):length(nms))
    cat(nms[i], "vs", nms[j], ":",
        length(intersect(gene_lists[[i]], gene_lists[[j]])), "\n")


```


# ==============================================================================
# STEP 5: Per-Dataset Volcano Plots (QC check before meta-analysis)
# Before combining datasets, visualise individual DE results to detectbatch effects, outliers, or platform-specific biases.

```{r}
plot_volcano <- function(de_df, title, fc_cut = 1, p_cut = 0.05) {
  de_df <- de_df %>%
    mutate(
      sig = case_when(
        abs(log2FC) >= fc_cut & pvalue < p_cut & log2FC  >  0 ~ "Up",
        abs(log2FC) >= fc_cut & pvalue < p_cut & log2FC  <  0 ~ "Down",
        TRUE ~ "NS"
      ),
      neg_log10_p = -log10(pvalue)
    )

  n_up   <- sum(de_df$sig == "Up",   na.rm = TRUE)
  n_down <- sum(de_df$sig == "Down", na.rm = TRUE)

  ggplot(de_df, aes(x = log2FC, y = neg_log10_p, colour = sig)) +
    geom_point(alpha = 0.5, size = 1.2) +
    scale_colour_manual(values = c(Up = "#E64B35", Down = "#4DBBD5", NS = "grey70")) +
    geom_vline(xintercept = c(-fc_cut, fc_cut), linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = -log10(p_cut),       linetype = "dashed", colour = "grey40") +
    labs(
      title    = title,
      subtitle = paste0("Up: ", n_up, "  |  Down: ", n_down),
      x        = expression(log[2]~"Fold Change"),
      y        = expression(-log[10]~"p-value"),
      colour   = "Direction"
    ) +
    theme_bw(base_size = 13) +
    theme(plot.title = element_text(face = "bold"))
}

for (gse_id in names(de_results)) {
  p <- plot_volcano(de_results[[gse_id]], title = paste(gse_id, "— Endometriosis vs Control"))
  ggsave(
    filename = file.path("results/plots", paste0(gse_id, "_volcano.pdf")),
    plot = p, width = 7, height = 6
  )
  message(gse_id, ": volcano plot saved")
}
```


#### STEP 6
# Run the three strategies
Random-Effects Model (recommended — use this as your primary result)
Models between-study heterogeneity explicitly. Most appropriate when datasets come from different platforms/cohorts, which is almost always the case with GEO.
```{r}
mv_rem <- rem_mv(
  diffexp       = mv_input,
  pcriteria     = "pvalue",
  foldchangecol = "log2FC",
  genenamecol   = "symbol",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  collaps       = FALSE   
)

write.csv(mv_rem@metaresult,
          "results/DE/MetaVolcanoR_REM_results.csv",
          row.names = FALSE)
message("rem_mv complete — ", nrow(mv_rem@metaresult), " genes")
```

The result slot mv_rem@metaresult gives you per-gene:
MeaningrandomSummary^:pooled log2FC estimate
randomP: meta p-value
tau2: between-study variance (heterogeneity)
I2: % of variance due to heterogeneity

SHow the plot
```{r}
mv_rem@MetaVolcano
```


Combining MV
```{r}
mv_fisher <- combining_mv(
  diffexp       = mv_input,
  pcriteria     = "pvalue",
  foldchangecol = "log2FC",
  genenamecol   = "symbol",
  collaps       = FALSE
)
```

Vote counting
```{r}
# ==============================================================================
mv_vote <- votecount_mv(
  diffexp       = mv_input,
  pcriteria     = "pvalue",
  foldchangecol = "log2FC",
  genenamecol   = "symbol",
  collaps       = FALSE
)
```


Get biological reuslts - Run fgsea
```{r}
library(fgsea)
library(msigdbr)
library(ggplot2)
library(dplyr)
library(forcats)

```

Download GO c5
```{r}
c5_sets <- msigdbr(species = "Homo sapiens", category = "C5")
pathways <- split(c5_sets$gene_symbol, c5_sets$gs_name)
cat("Total C5 gene sets:", length(pathways), "\n")

```


Build a gene list using the rem results
```{r}
rem_res <- mv_rem@metaresult

gene_ranks <- setNames(rem_res$randomSummary, rem_res$symbol)

```

Run fgsea
```{r}
set.seed(42)
fgsea_res <- fgsea(
  pathways  = pathways,
  stats     = gene_ranks,
  minSize   = 15,
  maxSize   = 500,
  nPermSimple = 10000
)

fgsea_res <- fgsea_res %>%
  arrange(pval) %>%
  filter(!is.na(padj))

cat("Significant pathways (padj < 0.05):", sum(fgsea_res$padj < 0.05, na.rm = TRUE), "\n")

```

Visualize the results
```{r}
top_n <- 30

plot_df <- fgsea_res %>%
  filter(padj < 0.01) %>%
  slice_min(padj, n = top_n) %>%
  mutate(
    direction = ifelse(NES > 0, "Up in endo", "Down in endo"),
    pathway_clean = gsub("^GOBP_|^GOCC_|^GOMF_", "", pathway),
    pathway_clean = gsub("_", " ", pathway_clean),
    pathway_clean = tolower(pathway_clean),
    pathway_clean = fct_reorder(pathway_clean, NES)
  )

lollipop <- ggplot(plot_df, aes(x = NES, y = pathway_clean, colour = direction)) +
  geom_segment(aes(x = 0, xend = NES, y = pathway_clean, yend = pathway_clean),
               linewidth = 0.7, alpha = 0.6) +
  geom_point(aes(size = -log10(padj)), alpha = 0.9) +
  scale_colour_manual(values = c("Up in endo" = "#C0392B", "Down in endo" = "#2980B9")) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  theme_minimal(base_size = 12) 

lollipop
```





