---
title: "MetaVolcanoR: Updated Features and Enrichment Analysis"
author: "Izabela Mamede, Bárbara Marinho"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MetaVolcanoR: Updated Features and Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(
  tidy    = FALSE,
  warning = FALSE,
  message = FALSE,
  cache   = TRUE,
  fig.width  = 8,
  fig.height = 5
)
```

## Overview

This vignette demonstrates the new features and improvements introduced
in this update of MetaVolcanoR:

- **Updated default color palette** for improved readability in 
  presentations and print
- **Fixed vote-counting axis** (`scale_x_continuous` replacing 
  `scale_x_discrete`)
- **Improved label selection** using the package's own ranking metrics 
  (`abs(idx)` for vote-counting, `rank` for REM)
- **Customizable DEG barplot** colors via new `colors` parameter in 
  `draw_degbar()`
- **`enrichment_mv()`**: new function for gene set enrichment analysis 
  with MSigDB integration and three ranking strategies
- **Protein-level meta-analysis**: MetaVolcanoR is feature-agnostic and 
  applies directly to differential protein abundance data (mass 
  spectrometry, Olink, SomaScan), demonstrated here on a vaccine 
  proteomics meta-analysis

## Setup

```{r packages}
library(MetaVolcanoR)
library(dplyr)
library(ggplot2)

data(diffexplist)
cat("Number of datasets:", length(diffexplist), "\n")
cat("Genes per dataset:", 
    paste(sapply(diffexplist, nrow), collapse = ", "), "\n")
```

## 1. Random Effects Model (REM)

The REM meta-analysis now uses an updated color palette and supports
pseudogene filtering in labels.

### Basic REM

```{r rem_basic}
mv_rem <- rem_mv(
  diffexp       = diffexplist,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  metathr       = 0.01,
  collaps       = TRUE,
  jobname       = "REM_example",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_rem@MetaVolcano
```

### REM with labels and pseudogene filtering

When `label_top_n` is used, genes are now ranked by the topconfects 
`rank` metric, which balances effect size and confidence.

```{r rem_labeled}
mv_rem_labeled <- rem_mv(
  diffexp       = diffexplist,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  metathr       = 0.01,
  collaps       = TRUE,
  label_top_n   = 10,
  label_size    = 3.5,
  plot_title    = "REM Meta-Analysis (Top 10)",
  show_legend   = TRUE,
  jobname       = "REM_labeled",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_rem_labeled@MetaVolcano
```

### Custom colors

The `colors` parameter accepts a named vector for the sign consistency
gradient:

```{r rem_custom_colors}
mv_rem_custom <- rem_mv(
  diffexp       = diffexplist,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  metathr       = 0.01,
  collaps       = TRUE,
  colors        = c(low = "purple", mid = "white", high = "orange", 
                    na = "gray80"),
  label_top_n   = 5,
  plot_title    = "REM with custom colors",
  jobname       = "REM_custom",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_rem_custom@MetaVolcano
```

## 2. Forest Plots

Forest plots now accept custom colors via a named vector:

```{r forest}
# Top 3 genes by rank
top3 <- mv_rem@metaresult %>%
  filter(!is.na(randomP)) %>%
  arrange(rank) %>%
  head(3) %>%
  pull(Symbol)

cat("Top 3 genes:", paste(top3, collapse = ", "), "\n")

# Forest plot with default updated colors
draw_forest(
  remres        = mv_rem,
  gene          = top3[1],
  genecol       = "Symbol",
  foldchangecol = "Log2FC",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  jobname       = "Forest_example",
  outputfolder  = tempdir(),
  draw          = "HTML"
)
```

Custom forest plot colors:

```{r forest_custom}
draw_forest(
  remres        = mv_rem,
  gene          = top3[1],
  genecol       = "Symbol",
  foldchangecol = "Log2FC",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  colors        = c(positive = "darkred", negative = "steelblue",
                    neutral = "gray70", reference = "black"),
  point_size    = 3,
  plot_title    = paste(top3[1], "- Custom colors"),
  jobname       = "Forest_custom",
  outputfolder  = tempdir(),
  draw          = "HTML"
)
```

## 3. Vote-Counting

### Fixed axis and improved labels

The vote-counting plot now uses `scale_x_continuous` for correct 
symmetric axis display. Labels are selected using `abs(idx)` — the 
package's own combined metric of frequency and direction consistency
(`idx = ndeg × ddeg`).

```{r vote}
mv_vote <- votecount_mv(
  diffexp       = diffexplist,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  geneidcol     = NULL,
  pvalue        = 0.05,
  foldchange    = 0,
  metathr       = 0.01,
  collaps       = TRUE,
  label_top_n   = 10,
  label_size    = 3.5,
  plot_title    = "Vote-Counting (Top 10 by |idx|)",
  jobname       = "Vote_example",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_vote@MetaVolcano
```

### Label selection criteria

Previous behavior ranked labels by `ndeg` only (number of studies where 
DE). This could label genes with inconsistent direction across studies.

The updated behavior uses `abs(idx)` and filters out unperturbed genes:

| Gene | ndeg | ddeg | idx | Previous rank | Updated rank |
|------|------|------|-----|---------------|--------------|
| A    | 5    | 5    | 25  | 1st (tied)    | 1st          |
| B    | 5    | -5   | -25 | 1st (tied)    | 2nd          |
| C    | 5    | 1    | 5   | 1st (tied)    | 3rd          |

### DEG barplot with custom colors

`draw_degbar()` now accepts a `colors` parameter:

```{r degbar}
mv_vote@degfreq
```

## 4. Combining (Fisher's Method)

Fisher's combining approach also supports the updated color palette:

```{r fisher}
mv_fisher <- combining_mv(
  diffexp       = diffexplist,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  metafc        = "Mean",
  metathr       = 0.01,
  collaps       = TRUE,
  colors        = c("#083e46", "gray90", "#811820"),
  point_size    = 1.2,
  show_legend   = TRUE,
  plot_title    = "Fisher's Combined Test",
  jobname       = "Fisher_example",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_fisher@MetaVolcano
```

## 5. Functional Enrichment — `enrichment_mv()`

The new `enrichment_mv()` function wraps fGSEA and integrates directly 
with the REM MetaVolcano results. Gene sets can be provided manually or 
downloaded automatically from MSigDB.

### Automatic MSigDB download

```{r enrich_auto, eval=FALSE}
# GO Biological Process (default: category = "C5")
enrich <- enrichment_mv(mv_rem, subcategory = "GO:BP")
enrich$plot
head(enrich$result)
```

### Three ranking strategies

```{r enrich_rankings, eval=FALSE}
# 1. Fold-change — prioritizes effect size
enrich_fc <- enrichment_mv(mv_rem, ranking = "fc")

# 2. Signed significance — prioritizes statistical confidence
enrich_sp <- enrichment_mv(mv_rem, ranking = "signed_p")

# 3. Weighted FC — balances both
enrich_wfc <- enrichment_mv(mv_rem, ranking = "weighted_fc")
```

| Ranking | Formula | Best for |
|---------|---------|----------|
| `fc` | randomSummary | Pathways with large effects |
| `signed_p` | -log10(p) × sign(FC) | Pathways with consistent significance |
| `weighted_fc` | FC × -log10(p) | Balanced detection |

### Selecting pathway databases

```{r enrich_databases, eval=FALSE}
# KEGG pathways
enrich_kegg <- enrichment_mv(
  mv_rem,
  category    = "C2",
  subcategory = "CP:KEGG",
  ranking     = "weighted_fc"
)

# Hallmark gene sets
enrich_hall <- enrichment_mv(
  mv_rem,
  category = "H"
)

# GO Molecular Function only
enrich_mf <- enrichment_mv(
  mv_rem,
  subcategory = "GO:MF"
)
```

### Full customization

```{r enrich_custom, eval=FALSE}
enrich <- enrichment_mv(
  mv_rem,
  category            = "C5",
  subcategory         = "GO:BP",
  ranking             = "weighted_fc",
  plot_padj           = 0.05,
  plot_top_n          = 20,
  clean_pathway_names = TRUE,
  colors              = c(down = "navy", up = "darkred"),
  plot_title          = "GO:BP Enrichment (weighted FC ranking)"
)

enrich$plot
```

### Using custom gene sets

```{r enrich_manual, eval=FALSE}
my_pathways <- list(
  "INFLAMMATION"  = c("IL6", "TNF", "IL1B", "CXCL8", "CCL2"),
  "ANGIOGENESIS"  = c("VEGFA", "FLT1", "KDR", "ANGPT1"),
  "ECM_REMODEL"   = c("MMP2", "MMP9", "MMP12", "COL1A1")
)

enrich_custom <- enrichment_mv(
  mv_rem,
  pathways            = my_pathways,
  clean_pathway_names = FALSE
)
```

## 6. Application to Proteomics Data

MetaVolcanoR is not limited to gene expression: every function operates on
any table of feature identifiers with a fold change, a p-value, and a
variance estimate. This means the same three meta-analysis strategies, the
same plots, and `enrichment_mv()` all apply directly to **differential
protein abundance** data — from mass-spectrometry proteomics (limma,
MSstats, DEqMS) or from affinity platforms such as Olink and SomaScan. The
only difference is what `genenamecol` identifiers mean: instead of gene
symbols from RNA-seq, they are gene-level protein symbols.

To demonstrate this, we meta-analyze the **day-7 versus day-0 plasma
proteomic response to seasonal influenza vaccination** across five
independent cohorts — a systems-vaccinology setting where a shared,
reproducible molecular signature is expected despite cohort-to-cohort
variation.

```{r prot_setup}
data(vaccine_proteomics)
cat("Number of cohorts:", length(vaccine_proteomics), "\n")
cat("Proteins per cohort:", 
    paste(sapply(vaccine_proteomics, nrow), collapse = ", "), "\n")
head(vaccine_proteomics[[1]])
```

Each cohort table has the same structure MetaVolcanoR expects for any
feature type: `Symbol` (here, a protein identifier), `Log2FC`, `pvalue`,
`CI.L`, `CI.R`. For proteomics results from limma/DEqMS, `CI.L`/`CI.R` are
typically derived from the standard error as `logFC ± 1.96 * SE`; the
`prepare_limma()` helper does this automatically.

### REM on proteins

```{r prot_rem}
mv_prot_rem <- rem_mv(
  diffexp       = vaccine_proteomics,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  metathr       = 0.01,
  collaps       = TRUE,
  label_genes   = c("IGHG1", "IGKC", "TNFRSF17", "MZB1", "CD38", "PRDM1"),
  label_size    = 3.5,
  plot_title    = "Day 7 vs Day 0 plasma proteome — influenza vaccine (5 cohorts)",
  jobname       = "VaccineProteome_REM",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_prot_rem@MetaVolcano
```

The proteins at the top of the ranking — immunoglobulin chains (`IGHG1`,
`IGKC`), the plasmablast marker `TNFRSF17` (BCMA), `CD38`, and the
plasma-cell factors `MZB1`, `PRDM1`, `XBP1` — recapitulate the canonical
day-7 antibody-secreting-cell signature of influenza vaccination, now
recovered at the **protein** level and reinforced by requiring consistency
across all five cohorts.

```{r prot_forest}
draw_forest(
  remres        = mv_prot_rem,
  gene          = "TNFRSF17",
  genecol       = "Symbol",
  foldchangecol = "Log2FC",
  llcol         = "CI.L",
  rlcol         = "CI.R",
  jobname       = "VaccineProteome_forest",
  outputfolder  = tempdir(),
  draw          = "HTML"
)
```

### Vote-counting and combining on proteins

```{r prot_vote}
mv_prot_vote <- votecount_mv(
  diffexp       = vaccine_proteomics,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  geneidcol     = NULL,
  pvalue        = 0.05,
  foldchange    = 0,
  metathr       = 0.01,
  collaps       = TRUE,
  label_top_n   = 10,
  label_size    = 3.5,
  plot_title    = "Vote-Counting: vaccine plasma proteins (Top 10 by |idx|)",
  jobname       = "VaccineProteome_vote",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_prot_vote@degfreq
mv_prot_vote@MetaVolcano
```

```{r prot_fisher}
mv_prot_fisher <- combining_mv(
  diffexp       = vaccine_proteomics,
  pcriteria     = "pvalue",
  foldchangecol = "Log2FC",
  genenamecol   = "Symbol",
  metafc        = "Mean",
  metathr       = 0.01,
  collaps       = TRUE,
  colors        = c("#083e46", "gray90", "#811820"),
  point_size    = 1.2,
  plot_title    = "Fisher's combined test — vaccine plasma proteome",
  jobname       = "VaccineProteome_fisher",
  outputfolder  = tempdir(),
  draw          = "HTML"
)

mv_prot_fisher@MetaVolcano
```

All three strategies converge on the same core of plasmablast- and
immunoglobulin-associated proteins, each perturbed in the same direction in
every cohort — the strongest evidence of reproducibility available in a
five-study design.

### Functional enrichment of the consistent proteins

Because `enrichment_mv()` operates on the REM result via gene-level
symbols, it works on the protein meta-analysis without modification.
Ranking the proteins by their summary statistics and running fGSEA against
MSigDB confirms that the consistent proteins are dominated by humoral /
B-cell and immunoglobulin programs — exactly the biology expected one week
after vaccination.

```{r prot_enrich, eval=FALSE}
# GO Biological Process, balanced ranking
enrich_prot_bp <- enrichment_mv(
  mv_prot_rem,
  category    = "C5",
  subcategory = "GO:BP",
  ranking     = "weighted_fc",
  plot_padj   = 0.05,
  plot_top_n  = 20,
  plot_title  = "GO:BP enrichment — vaccine plasma proteome"
)
enrich_prot_bp$plot
head(enrich_prot_bp$result)

# Immunologic signature collection (C7) is especially apt for vaccine data
enrich_prot_c7 <- enrichment_mv(mv_prot_rem, category = "C7", ranking = "signed_p")
enrich_prot_c7$plot
```

Expected top pathways include immunoglobulin-mediated immune response,
B-cell / plasma-cell differentiation, and humoral response terms — a
functional readout that ties the protein-level meta-analysis back to the
mechanisms systems vaccinology has long characterized at the
transcriptional level.

## Summary of changes

| Feature | Before | After |
|---------|--------|-------|
| Default palette | blue/grey/red | #083e46/grey90/#811820 |
| Vote-counting axis | `scale_x_discrete` (truncated) | `scale_x_continuous` (symmetric) |
| Vote-counting labels | ranked by `ndeg` | ranked by `abs(idx)` |
| REM labels | ranked by `randomP` | ranked by `rank` (topconfects) |
| DEG barplot colors | hardcoded | `colors` parameter |
| Enrichment analysis | not available | `enrichment_mv()` |
| Protein-level data | not demonstrated | full worked example (vaccine proteomics, §6) |
